---
title: "Data sources, species and countries"
format: 
  html:
    toc: true
    toc-location: right
    smooth-scroll: true
    html-math-method: katex
    df-print: kable
    code-fold: true
author: 'Florencia Grattarola'
date: "`r format(Sys.time(), '%Y-%m-%d')`"
editor: source
self-contained: true
---

R libraries

```{r}
#| label: libraries
#| echo: true
#| eval: true
#| message: false
#| warning: false

library(tmap)
tmap_mode('view')
library(sf)
sf::sf_use_s2(FALSE)
library(tidyverse)
```


### List of species

```{r}
#| label: species
#| echo: true
#| eval: true
#| message: false
#| warning: false

carnivores <- read_csv('data/carnivores.csv')
carnivores %>% kableExtra::kbl()
```

**Number of carnivore species** = `r nrow(carnivores)`

### Countries

```{r}
#| label: countries
#| echo: true
#| eval: true
#| message: false
#| warning: false

countries <- st_read('data/latam.gpkg', layer = 'countries', quiet = T)
countries_land <- st_read('data/latam.gpkg', layer = 'countries_land', quiet = T)

tm_shape(countries_land) +
    tm_fill(col = 'name_en',
            palette = 'Set1', 
            title= '') + 
    tm_layout(legend.outside = T) 
```


### Data sources

```{r}
#| label: data
#| echo: true
#| eval: true
#| message: false
#| warning: false

PO <- read_csv('data/data_PO.csv')
PA <- read_csv('data/data_PA.csv')

PO %>% summarise(`Number of records`=n(),
                 `Number of species`=n_distinct(scientificName),
                 `Number of datasets`=n_distinct(datasetKey)) %>% 
  kableExtra::kbl(caption = 'presence-only', align = 'l')

PA %>% summarise(`Number of records` = n(),
                 `Number of records`= n(),
                 `Number of species` = n_distinct(scientificName),
                 `Number of sources` = n_distinct(datasetName)) %>% 
  kableExtra::kbl(caption = 'presence-absence', align = 'l')
```

**Total number of species covered** = `r nrow(bind_rows(PO %>% distinct(scientificName), PA%>% distinct(scientificName)) %>% distinct(scientificName))`. 

#### Species not covered

```{r}
#| label: species-not-covered
#| echo: true
#| eval: true
#| message: false
#| warning: false

carnivores %>% 
  filter(!species %in%
           (bind_rows(PO %>% distinct(scientificName),
                      PA %>% distinct(scientificName)) %>% 
              pull(scientificName))) %>%
  select(species, family) %>% 
  kableExtra::kbl()
```

#### Number of records per species

```{r}
#| label: data-count-per-species
#| echo: true
#| eval: true
#| message: false
#| warning: false

PO %>% count(scientificName, family) %>% 
  arrange(family, scientificName) %>% 
  kableExtra::kbl(caption = 'presence-only', align = 'l')

PA %>% count(scientificName, family) %>% 
  arrange(family, scientificName) %>% 
  kableExtra::kbl(caption = 'presence-absence', align = 'l')
```

### Files 

Presence-only column names

```{r}
#| label: data-columns-PO
#| echo: true
#| eval: true
#| message: false
#| warning: false

PO %>% names() %>% 
  as_tibble_col(column_name = 'column name') %>% 
  kableExtra::kbl(align = 'l')
```


Presence-absence column names

```{r}
#| label: data-columns-PA
#| echo: true
#| eval: true
#| message: false
#| warning: false

PA %>% names() %>% 
  as_tibble_col(column_name = 'column name') %>% 
  kableExtra::kbl(align = 'l')
```

### Save the metadata files

```{r}
#| label: PA-polygons-save
#| echo: true
#| eval: false
#| message: false
#| warning: false
#| code-fold: false

PO %>% names() %>%
  as_tibble_col(column_name = 'columnName') %>%
  write_excel_csv('metadata/PO_metadata.csv')

PA %>% names() %>%
  as_tibble_col(column_name = 'columnName') %>%
  write_excel_csv('metadata/PA_metadata.csv')
```